Reproducible Research - Assignment 1
========================================================

### Load required packages
```{r, echo=TRUE}
library(dplyr)
library(ggplot2)
library(lattice)
```

### Load Dataset 
The dataset should be available in the working directory
```{r, echo=TRUE}
path <- file.path(getwd(),"activity.csv")
activity <- read.csv(path)
```

### Checking Data
Get an overview of the data
```{r,echo=TRUE}
str(activity)
names(activity)
head(activity, 20)
tail(activity, 20)
```

# What is the mean of total number of steps taken per day?
### Manipulating the data
Calculate sum, mean and median for each day
```{r, echo=TRUE}
activity_1 <- activity %>% group_by(date) %>% 
        summarise(sum_day = sum(steps, na.rm=TRUE),
                  mean_day = mean(steps, na.rm=TRUE),
                  median_day = median(steps, na.rm=TRUE))
```

### Histogram of total number of steps taken each day
```{r, echo=TRUE}
hist(activity_1$sum_day, col="blue", breaks=20, 
     main="Frequency of Total Steps Per Day", xlab="Steps/day")
rug(activity_1$sum_day)
```
### Report of mean and median of the total number of steps taken each day
```{r, echo=TRUE}
print(tbl_df(activity_1[,3:4]), n=61)
```

# What is the average daily pattern?
### Manipulating the data
Calculate the mean of the intervals
```{r, echo=TRUE}
activity_2 <- activity %>% group_by(interval) %>%
        summarise(mean_int = mean(steps, na.rm=TRUE))
```
### Show the maximum mean at the interval
```{r, echo=TRUE}
filter(activity_2, mean_int == max(mean_int))
```

### Creating the time series plot
```{r, echo=TRUE}
p <- ggplot(data=activity_2, aes(x=interval, y=mean_int))
p <- p + geom_line(size=1, color="red")
p <- p + theme_bw()
p <- p + ggtitle("Time series plot of the 5-minute interval and \nthe average steps taken in the interval")
p <- p + xlab("5-minute interval")
p <- p + ylab("Mean of steps")
p
```

# Input missing values 

### Calculate the total number of rows with NAs
```{r, echo=TRUE}
activity_complete <- complete.cases(activity)
activity_3 <- activity[activity_complete,]
nrow(activity)-nrow(activity_3)
```

### Strategy for filling missing values
Create a new file
```{r, echo=TRUE}
activity_4 <- activity
```

I decided to calculate the mean of interval for the missing values
```{r, echo=TRUE}
for(i in 1:nrow(activity_4)){
        if(is.na(activity_4[i,1])){
                activity_4[i,1] <- activity_4 %>% 
                        filter(interval == activity_4[i,3]) %>%
                        summarise(mean(steps, na.rm=TRUE))}}
```

### Calculation sum, mean and median - no NAs
```{r, echo=TRUE}
activity_5 <- activity_4 %>% group_by(date) %>% 
        summarise(sum_day = sum(steps, na.rm=TRUE),
                  mean_day = mean(steps, na.rm=TRUE),
                  median_day = median(steps, na.rm=TRUE))
```

### Histogram of total number of steps taken each day - no NAs
```{r, echo=TRUE}
hist(activity_5$sum_day, col="blue", breaks=20,
     main="Frequency of Total Steps Per Day - no NAs", xlab="Steps/day",
     ylim=c(0,20))
rug(activity_5$sum_day)
```

Just to show the difference between the two vectors (steps)
```{r, echo=TRUE}
activity_5$sum_day - activity_1$sum_day
```
There is an increase of frequency (8) by 10766.19

# Are there differences in activity patterns between weekdays and weekends?
Work with the new dataset (no NAs)
```{r, echo=TRUE}
activity_6 <- activity_4
```

### Convert date from factor to Date
```{r, echo=TRUE}
activity_6$date <- as.Date(activity_6$date)
```

### Generate a new factor variable
Please note that I come from germany and use the german words for saturday (Samstag) and sunday (Sonntag)
```{r, echo=TRUE}
activity_6$daytype <-as.factor(ifelse(weekdays(activity_6$date) == "Samstag" |
                                    weekdays(activity_6$date) == "Sonntag",
                            "Weekend", "Weekday"))
```

### Manipulating the data
Calculate the mean of the interval for each day (with the new dataset)
```{r, echo=TRUE}
activity_7 <- activity_6 %>% group_by(interval, daytype) %>%
        summarise(sum_day=mean(steps))
```

### Create a panel plot 
```{r, echo=TRUE}
p <- ggplot(data = activity_7, aes(x=interval,y=sum_day, group=daytype))
p <- p + geom_line(size=1,color="green")
p <- p + theme_bw()
p <- p + facet_grid(. ~ daytype,scales="free_y")
p <- p + ggtitle("Comparision of 5-minute interval on weekdays and weekends")
p <- p + xlab("5-minute interval")
p <- p + ylab("Mean of steps")
p
```

As can be seen in the panel plot the average amount of steps rises faster during
weekdays (~interval 500) and higher (~ interval between 500 and 1000) than on
weekends. On weekends the average amount of steps rises more gradually
(~ interval 500 - 1000).
